BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2015-03-12 Paper/D5.5/D5.5.tex
Added results for Laplace and BigDFT.

\section{Motivating Example: OpenCL Laplace}

During one of the Mont-Blac face to face meeting a talk on OpenCL optimization
on the Mali GPU was given. One of the case study was a Laplace filter on the
GPU~\cite{opencl_arm_training}.

@@

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized with Synthesized Loads},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_vector_synth}}
\begin{lstlisting}
  uchar16 v11_ = vload16( 0, src + 3*width*(j-1) + 3*5*i - 3 );
  uchar16 v13_ = vload16( 0, src + 3*width*(j-1) + 3*5*i + 3 );
  uchar16 v12_ = uchar16( v11_.s3456789a, v13_.s56789abc );
  uchar16 v21_ = vload16( 0, src + 3*width*(j  ) + 3*5*i - 3 );
  uchar16 v23_ = vload16( 0, src + 3*width*(j  ) + 3*5*i + 3 );
  uchar16 v22_ = vload16( v21_.s3456789a, v23_.s56789abc );
  uchar16 v31_ = vload16( 0, src + 3*width*(j+1) + 3*5*i - 3 );
  uchar16 v33_ = vload16( 0, src + 3*width*(j+1) + 3*5*i + 3 );
  uchar16 v32_ = vload16( v31_.s3456789a, v33_.s56789abc );
\end{lstlisting}

@@

\begin{lstlisting}
void magicfilter(const int32_t n, const int32_t ndat,
                 const double * x, double * y)
}
\end{lstlisting}
\caption{BOAST Code Snippet for Procedure}
\label{fig:BOAST_procedure}
\end{figure}

@@

In this section we will present two scientific applications that use BOAST. The
first one, BigDFT, uses boast in order to develop new functionalities with
performance portability in mind. The second one, SPECFEM3D, uses BOAST to
factorize OpenCL and CUDA development, while having robust non regression
tests.

@@

The interest in having a robuts and optimally tuned library in this scientific
field is therefore evident.  Techniques are under investigation to provide also
fine-tuned binaries to the end-user rather than the code sources, such that
more agressive interprocedural optimization can be performed.  This experience
would be therefore a first step toward the release of a tunable optimised
convolution library oriented to computational physics communities.
%   BigDFT implementation of convolution/wavelets is a reference but is limited
%   in scope (1 wavelet family ands its derived operators).  Generalizing to
%   other wavelet families long and difficult if performance are to remain
%   good.
% 
% Idea: \begin{itemize} \item write a generic convolution library in BOAST
% \item BLAS like interface \item Provide a wide variety of building blocks
% \item Optimize thos blocks using BOAST \item Build either from source or
% directly link with binary \item Release complete library \end{itemize}
% 
% 
%   Characteristic use case driven
% 
%   Convolution Library: first step toward a dsl dedicated to wavelet
%   processing for Physics computing.

>>>>

\section{Motivating Example: OpenCL Laplace}
\label{sec:laplace}

During one of the Mont-Blanc face to face meeting a talk on OpenCL optimization
on the Mali GPU was given. One of the case study was a Laplace filter on the
GPU~\cite{opencl_arm_training}.

@@

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized with Synthesized Loads},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_vector_synth}}
\begin{lstlisting}
  uchar16 v11_ = vload16( 0, src + 3*width*(j-1) + 3*5*i - 3 );
  uchar16 v13_ = vload16( 0, src + 3*width*(j-1) + 3*5*i + 3 );
  uchar16 v12_ = uchar16( v11_.s3456789a, v13_.s56789abc );
  uchar16 v21_ = vload16( 0, src + 3*width*(j  ) + 3*5*i - 3 );
  uchar16 v23_ = vload16( 0, src + 3*width*(j  ) + 3*5*i + 3 );
  uchar16 v22_ = uchar16( v21_.s3456789a, v23_.s56789abc );
  uchar16 v31_ = vload16( 0, src + 3*width*(j+1) + 3*5*i - 3 );
  uchar16 v33_ = vload16( 0, src + 3*width*(j+1) + 3*5*i + 3 );
  uchar16 v32_ = uchar16( v31_.s3456789a, v33_.s56789abc );
\end{lstlisting}

@@

\begin{lstlisting}
void magicfilter(const int32_t n, const int32_t ndat,
                 const double * x, double * y) {
}
\end{lstlisting}
\caption{BOAST Code Snippet for Procedure}
\label{fig:BOAST_procedure}
\end{figure}

@@

In this section we will present the benefits of using BOAST on the Laplace
motivating example as well as two scientific applications that use BOAST. The
first one, BigDFT, uses boast in order to develop new functionalities with
performance portability in mind. The second one, SPECFEM3D, uses BOAST to
factorize OpenCL and CUDA development, while having robust non regression
tests.

  \subsection{Laplace Filter Kernel}

Section~\ref{sec:laplace} presented the Laplace motivating example. From this
section we know that a number of optimization can have an impact on the
performance of the kernel on the Mali architecture. But what of other
architectures? And are there other optimizations that can impact the
performance?

    \subsubsection{Optimization Space}

The list of already identified optimizations are:
\begin{itemize}
\item vectorization,
\item intermediary data type,
\item number of pixels processed,
\item and synthesizing loads.
\end{itemize}
To create our generic implementation we decided to work at the component level
rather than the pixel level. This approach leads to more flexibility and
genericity when applying optimizations. We also decided to study the impact of
another parameter which is the number of component to process on the column
direction. This leads to being able to process tiles instead of only rows.

Here are the parameters we finally selected for our kernel optimization and
their possible values:
\begin{itemize}
\item \textit{x\_component\_number}: a positive integer
\item \textit{y\_component\_number}: a positive integer
\item \textit{vector\_length}: 1, 2, 4, 8 or 16
\item \textit{temporary\_size}: 2 or 4
\item \textit{synthesize\_loads}: \textit{true} or \textit{false}
\item \textit{vector\_recompute}: \textit{true} or \textit{false}
\end{itemize}
The last parameter is used when \textit{x\_component\_number} is not divisible by
\textit{vector\_length}. Two solutions are possible then, divide the remainder
of the division in vectors of smaller sizes (\textit{vector\_recompute} =
\textit{false}) or load more data and compute useless values in vectors of the
specified size. This last option is mimicking the behavior of ARM
implementation, although when working at the component level it may seem an
absurd thing to try.

    \subsubsection{Performance Results}

Table~\ref{tbl:ARM} are the best results obtained by ARM on different image
compared to the naive implementation and to the best version version BOAST
found. It shows that the generated version systematically beats the hand
optimized version. As far as the optimization options are concerned the result
are a disappointment: the same kernel configuration is the best for all image
sizes. This kernel is using \textit{x\_component\_number} = 16,
\textit{y\_component\_number} = 1, \textit{vector\_length} = 16,
\textit{temporary\_size} = 2 and \textit{synthesize\_loads} = false.
\textit{vector\_recompute} has no meaning here as \textit{x\_component\_number}
== \textit{vector\_length}. Those results also shows that, when working on full
vectors, synthesizing the loads is harmful to performance and the programmer is
better of trusting the cache to load each vector in 1 cycle without compromising
the bandwidth.

\begin{table}
\centering
\begin{tabular}{|l|r|r|r|r|r|}
\hline
Image Size  & Naive (s) & Best (s) & Acceleration & BOAST (s) & Acceleration \\
\hline
768 x 432   & 0.0107    & 0.00669  & x1.6         & 0.000639  & x16.7 \\
\hline
2560 x 1600 & 0.0850    & 0.0137   & x6.2         & 0.00687   & x12.4 \\
\hline
2048 x 2048 & 0.0865    & 0.0149   & x5.8         & 0.00715   & x12.1 \\
\hline
5760 x 3240 & 0.382     & 0.0449   & x8.5         & 0.0325    & x11.8 \\
\hline
7680 x 4320 & 0.680     & 0.0747   & x9.1         & 0.0581    & x11.7 \\
\hline
\end{tabular}
\caption{Best performance of ARM Laplace kernel.}
\label{tbl:ARM}
\end{table}

But what if we run out benchmark on other architectures? Table~\ref{tbl:other}
the results obtained when running our BOAST implementation on other
architectures. The chosen architectures were an Intel i7-2760QM CPU (Sandy
Bridge architecture) that supports OpenCL 1.2 and an NVIDIA gtx680 GPU that
support OpenCL 1.1. We can see that the performance ratio between the different
architectures is rather stable across image sizes.

The optimization parameters selected are not the same for those architectures.
Indeed, the Intel CPU favors kernels that have the parameters:
\textit{x\_component\_number} = 16,
\textit{vector\_length} = 8,
\textit{temporary\_size} = 2 and \textit{synthesize\_loads} = false. Once again
\textit{vector\_recompute} has no meaning. \textit{y\_component\_number} varies
from 4 to 2 when image size increase, thus decreasing task granularity as the
global work size increase. This result is unexpected and understanding why this
happen could be interesting. The NVIDIA GPU favors processing square tiles:
\textit{x\_component\_number} = 4, \textit{y\_component\_number} = 4,
\textit{vector\_length} = 4, \textit{temporary\_size} = 2 and
\textit{synthesize\_loads} = false. Once more \textit{vector\_recompute} has no
meaning.

\begin{table}
\centering
\begin{tabular}{|l|r|r|r|r|r|}
\hline
Image Size & BOAST ARM (s) & BOAST Intel & Ratio & BOAST NVIDIA & Ratio \\
\hline
768 x 432   & 0.000639     & 0.000222    & x2.9  & 0.0000715    & x8.9  \\
\hline
2560 x 1600 & 0.00687      & 0.00222     & x3.1  & 0.000782     & x8.8  \\
\hline
2048 x 2048 & 0.00715      & 0.00226     & x3.2  & 0.000799     & x8.9  \\
\hline
5760 x 3240 & 0.0325       & 0.0108      & x3.0  & 0.00351      & x9.3  \\
\hline
7680 x 4320 & 0.0581       & 0.0192      & x3.0  & 0.00623      & x9.3  \\
\hline
\end{tabular}
\caption{Best performance of Laplace Kernel on several architectures.}
\label{tbl:other}
\end{table}

    \subsubsection{Laplace Conclusion}

In this Subsection we have shown the interest of BOAST in optimizing a well
known algorithm across different architectures. Chosen optimization combinations
are highly dependant on the targeted architecture. In the next Subsections we
will show that our methodology also applies to real applications.

@@

The interest in having a robuts and optimally tuned library in this scientific
field is therefore evident.  Techniques are under investigation to provide also
fine-tuned binaries to the end-user rather than the code sources, such that
more agressive interprocedural optimization can be performed.  This experience
would be therefore a first step toward the release of a tunable optimised
convolution library oriented to computational physics communities.

    \subsubsection{Performance report}

Several kernels have already been implemented in BOAST for the convolution
library. Figure~\ref{fig:synthesis} show the performance of the wavelet
transform operation as a factor of the unrolling length of the outer loop, the
language used to implement it as well as the activation or not of the OpenMP
parallelization. Results are given as a speedup compared to the sequential hand
tuned implementation that can be found in BigDFT. These tests were run on the
Inte Xeon X5550 that was used to hand optimize the code.

We can see that this function is better optimized using FORTRAN and small
unrolling factors. In the hand optimized version the unrolling factor was chosen
much too high (a factor of 12 was used). This factor might have been optimal at
the point the procedure was optimized (compiler version changed in the meantime) but
since then the environment changed. Other optimization have also been
incorporated in the BAOST sources, like the systematic inner loop unrolling, and
those could also help increase performance while limiting the interest of the
outer loop unrolling.

Nonetheless what is interesting from the physicist point of view is that the
generated source will give her better performance on a whole range of
architectures/compiler combination than that of the hand tuned code.

\begin{figure}
\begin{center}
\includegraphics[scale=0.7]{Res_synthesis}
\caption{Impact of Unrolling, Language and OpenMP on a Wavelet Transform}
\end{center}
\label{fig:synthesis}
\end{figure}
%   BigDFT implementation of convolution/wavelets is a reference but is limited
%   in scope (1 wavelet family ands its derived operators).  Generalizing to
%   other wavelet families long and difficult if performance are to remain
%   good.
% 
% Idea: \begin{itemize} \item write a generic convolution library in BOAST
% \item BLAS like interface \item Provide a wide variety of building blocks
% \item Optimize thos blocks using BOAST \item Build either from source or
% directly link with binary \item Release complete library \end{itemize}
% 
% 
%   Characteristic use case driven
% 
%   Convolution Library: first step toward a dsl dedicated to wavelet
%   processing for Physics computing.
